---
title: "Results"
format: html
---

```{r , include=FALSE}
# R 4.4.1 | Created by Brady Halvorson with River Focus, Inc. November 8, 2024.
# R 4.4.2 | Prepared for integration in Quarto by Benjamin Meyer, Kenai Watershed Forum, December 17, 2024

knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)

# clear environment
# rm(list=ls())

# load packages
library(tinytex)
library(tidyverse)
library(knitr)
library(xfun)
library(openxlsx)
library(openxlsx2)
library(plotly)
library(readxl)
library(htmlwidgets)
```

## Ranking by Site

We plotted each site by it's physical and ecological scores on XY axes (@fig-quadrant). Each point represents an individual site. Click each point to see it's location in Google Maps in a new web browser page.

Note that points in @fig-quadrant are jittered, meaning that they are deliberately spread out slighty in order to visualize values that would otherwise be superimposed atop each other.

```{r fig-quadrant, fig.cap="Test"}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-label: fig-quadrant

# Export and prep

# read in merged data set
today <- format(Sys.Date(), "%m%d%Y") # MMDDYYYY format
master_data_set_file_name <- paste0("other/output/","CombinedData_", today, ".xlsx")
merged_data_set <- read_excel(master_data_set_file_name)

# Prep master data sheet with coordinates hyperlinked to Google Maps
merged_data_set <- merged_data_set |>
  mutate(location_hyperlink = paste0("https://www.google.com/maps/search/?api=1&query= ",Longitude,Latitude))


# Create the plot
#p <- ggplot(merged_data_set, aes(x = Priority_Ecological, y = Priority_Physical,
#                                 text = paste("Location:",Latitude,Longitude,"\n",
#                                             "Stream:",`Stream Name`,"\n",
#                                              "Google Maps", location_hyperlink))) +
#  geom_jitter(color = "blue", size = 2, width = .4, height = 0.4) +
#  labs(x = "Ecological Condition", y = "Physical Condition") +
#  theme_minimal() +


  # Add the priority zones (adjust colors and labels as needed)
  #geom_rect(xmin = 0, xmax = 10, ymin = 0, ymax = 5, fill = "pink", alpha = 0.3) +
  #geom_rect(xmin = 10, xmax = 20, ymin = 0, ymax = 5, fill = "#199e19", alpha = 0.2) +
  #geom_rect(xmin = 0, xmax = 10, ymin = 5, ymax = 10, fill = "lightblue", alpha = 0.2) +
  #geom_rect(xmin = 10, xmax = 20, ymin = 5, ymax = 10, fill = "yellow", alpha = 0.2) + 
  
  # Add priority labels
#  annotate("text", x = 5, y = 2.5, label = "4 = Lowest Priority", color = "black") +
#  annotate("text", x = 15, y = 2.5, label = "2 = High-Middle Priority", color = "black") +
#  annotate("text", x = 5, y = 7.5, label = "3 = Low-Middle Priority", color = "black") +
#  annotate("text", x = 15, y = 7.5, label = "1 = Highest Priority", color = "black") +

  # Add axes labels
#  geom_segment(aes(x = -.5, y = -.5, xend = 20, yend = -.5)) +
#  geom_segment(aes(x = -.5, y = -.5, xend = -.5, yend = 12)) +
#  geom_segment(aes(x = 10, y = -.5, xend = 10, yend = 12), linetype = 2) +
#  geom_segment(aes(x = -.5, y = 6, xend = 20, yend = 6), linetype = 2)
  

# make plotly
#ggplotly(p)


# cleate plotly object


# prep plotly graph object
# modified from River Focus draft by Benjamin Meyer 12/30/2024
p <- plot_ly(merged_data_set, 
             x = ~jitter(Priority_Ecological), 
             y = ~jitter(Priority_Physical)) %>%
  layout(xaxis = list(zeroline = FALSE, title = "Ecological Priority"), 
         yaxis = list(zeroline = FALSE, title = "Physical Priority"),
         # annotations
         annotations = list(
           list(x = 5, y = 2.5, text = "4 = Lowest Priority", color = "black", showarrow = FALSE),
           list(x = 15, y = 2.5, text = "2 = High-Middle Priority", color = "black", showarrow = FALSE),
           list(x = 5, y = 7.5, text = "3 = Low-Middle Priority", color = "black", showarrow = FALSE),
           list(x = 15, y = 7.5, text = "1 = Highest Priority", color = "black", showarrow = FALSE)),
         showlegend = FALSE) %>%
  # interactive points
  add_markers(
    text = paste("Location:",merged_data_set$Latitude,merged_data_set$Longitude,"\n",
                 "Stream:",merged_data_set$`Stream Name`,"\n",
                 "Click to Open Map in Web Browser"),
    customdata = paste0("https://www.google.com/maps/search/?api=1&query= ",
                        merged_data_set$Latitude, ", ", merged_data_set$Longitude)) %>%
  # line segments
  add_segments(x = -.5,  y = -.5, xend = 20, yend = -.5, line = list(color = "black", width = 2)) %>%
  add_segments(x = -.5,  y = -.5, xend = -.5, yend = 12, line = list(color = "black", width = 2)) %>%
  add_segments(x = 10,  y = -.5, xend = 10, yend = 12, line = list(color = "black", width = 1, dash = "dash")) %>%
  add_segments(x = -.5,  y = 6, xend = 20, yend = 6, line = list(color = "black", width = 1, dash = "dash"))


# render plotly object with points hyperlinked 
onRender(
  p, "
  function(el) {
    el.on('plotly_click', function(d) {
      var url = d.points[0].customdata;
      window.open(url);
    });
  }
")

```



<br>

## Online Project Map

The project map may be accessed by following the link at [ArcGIS Online](https://kwf.maps.arcgis.com/apps/mapviewer/index.html?webmap=c61b78fa1cd24726ab90b40308f932d6)[^results-1] or in the interactive figure below. Toggle layers on/off as needed by clicking the "Layers" icon on the left hand side of the map.

[^results-1]: <https://kwf.maps.arcgis.com/apps/mapviewer/index.html?webmap=c61b78fa1cd24726ab90b40308f932d6>

<iframe width="800" height="600" frameborder="0" scrolling="no" allowfullscreen src="https://kwf.maps.arcgis.com/apps/mapviewer/index.html?webmap=c61b78fa1cd24726ab90b40308f932d6">

</iframe>

</iframe>

<br>
