# Summary

```{r , include=FALSE}
# R 4.4.1 | Created by Brady Halvorson with River Focus, Inc. November 8, 2024.
# R 4.4.2 | Prepared for integration in Quarto by Benjamin Meyer, Kenai Watershed Forum, December 17, 2024

knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)

# clear environment
# rm(list=ls())

# load packages
library(tinytex)
library(tidyverse)
library(knitr)
library(xfun)
library(openxlsx)
library(plotly)
library(readxl)
```

-   Most fish passage structures inventoried do not have

-   Our method is biased to rank sites higher where more site details about the physical infrastructure are available. A

-   Eco - based on known hab only. does not include lakes. current rankings a reflection of where feasible to do fieldwork,

-   Additional fieldwork to update these datasets

-   Next steps

    -   Main summarized steps

    -   link to downloadable document with more detailed recs

-   

```{r}

#| code-fold: true
#| code-summary: "Show the code"

# read in merged data set
today <- format(Sys.Date(), "%m%d%Y") # MMDDYYYY format
master_data_set_file_name <- paste0("other/output/","CombinedData_", today, ".xlsx")
merged_data_set <- read_excel(master_data_set_file_name)

tbl <- merged_data_set |>
  group_by(Perch_Height,Constriction_Ratio,Structure_Slope) |>
  count()

tbl

```
