# Summary

```{r , include=FALSE}
# R 4.4.1 | Created by Brady Halvorson with River Focus, Inc. November 8, 2024.
# R 4.4.2 | Prepared for integration in Quarto by Benjamin Meyer, Kenai Watershed Forum, December 17, 2024

knitr::opts_chunk$set(echo = TRUE, warning = F, message = F)

# clear environment
# rm(list=ls())

# load packages
library(tinytex)
library(tidyverse)
library(knitr)
library(xfun)
library(openxlsx)
library(plotly)
library(readxl)
library(DT)
library(htmltools)
```

Our results constitute a published draft method that can annually integrate multiple, distinct data inventories of fish passage structures (e.g. culverts) and freshwater habitat within the Kenai Peninsula Borough, then apply that data to calculate a relative prioritization score.

The regional prioritization scores provide a useful and important perspective to help managers decide how and when restoration funds should be directed. However, before considering this exercise to be complete, several additional needs must be resolved, which are described below (as of `r Sys.Date()`):

1.  **Address missing culvert physical descriptor data:** In the final combined data set, most fish passage structures inventoried do not have any associated physical data (\~350 out of 412 sites). See the table below, which displays unique row counts for all locations in the final data set. The "row_count" column indicates the number of instances with unique combinations of perch high, constriction ratio, and slope.

    -   It is likely that the low number of sites with documented culvert structure conditions in the final data set is an artifact of the code synthesizing these data sets. Some column descriptors may have been dropped when they were incompatible among agencies. A solution is likely achieved by revising how Alaska Dept of Fish and Game and US Fish and Wildlife Service data sets overlap.

    -   The prioritization algorithm described in this report can not currently incorporate physical descriptors data, as it is largely not present in the final combined dataset.

2.  **Incorporate probable upstream anadromous habitat**: The current final combined dataset includes reported (e.g documented in the ADF&G Anadromous Waters Catalog, or AWC) stream miles above each fish passage structure. The current final combined dataset is incomplete in two ways:

    -   The ADF&G AWC is estimated to include at most \~50% of all actual anadromous water bodies. Water bodies are presumed to be *not* anadromous unless otherwise documented. Documenting and modeling anadromous waters is a topic of active research in Alaska and the Pacific Northwest[^summary-1],[^summary-2].

    -   The initial work plan for activities in this report included use of probable upstream anadromous habitat. Kenai Watershed Forum generated a modified shape file of the National Hydrography Database, which estimated the total extent of probable anadromous habitat based on the locations of potential barriers to upstream migration of adult salmon. However, during this project it was discovered that this shape file required multiple additional QA/QC steps, and addressing them was beyond the scope of current efforts.

    -   The current algorithm incorporates upstream stream miles, but does not include lake acres.

3.  Document all locations in the KPB where streams intersect with with transportation and utility corridor infrastructure

    -   Perform a desktop GIS exercise intersecting borough transportation and utility corridor shape files (roads, trails, rail, utility corridors) with the National Hydrography Databse. Create a shape file of point locations from this intersection. (Results from this exercise may be available from the Palmer Soil and Water Conservation District office).
    -   

[^summary-1]: <https://app.paperpile.com/view/plain/?id=ba5fcffa-9dec-4171-8b81-68831d4e99f9>

[^summary-2]: <https://www.kenaiwatershed.org/about-kwf/news-media/>

```{r, combined-table}

#| code-fold: true
#| code-summary: "Show the code"

# read in merged data set
today <- format(Sys.Date(), "%m%d%Y") # MMDDYYYY format
master_data_set_file_name <- paste0("other/output/","CombinedData_", today, ".xlsx")
merged_data_set <- read_excel(master_data_set_file_name)

tbl <- merged_data_set |>
  group_by(Perch_Height,Constriction_Ratio,Structure_Slope) |>
  count() |>
  rename(row_count = n)

tbl1 <- tbl |>
  arrange(desc(row_count))

# publish table
datatable(tbl1, options = list(pageLength = 5),
          caption = "Table 1. Unique row count for culvert physical conditions data; combined Kenai Peninsula Borough fish passage datasets. (Version 1.0, 12/23/2024)") 
  


```

-   Our method is biased to rank sites higher where more site details about the physical infrastructure are available. A

-   Eco - based on known hab only. does not include lakes. current rankings a reflection of where feasible to do fieldwork,

-   Additional fieldwork to update these datasets

-   Next steps

    -   Main summarized steps

    -   link to downloadable document with more detailed recs

-   

-   Our results constitute a proof of concept and draft in progress. Additional fieldwork and analysis efforts are needed to ensure complete information, including:

    -   

    -   test
